---
title: "rmarkdown example"
author: "MdM"
date: "March 3, 2019"
output: html_document
---


```{r setup, include=FALSE} 
knitr::opts_chunk$set(echo = TRUE)
# This is the standard .Rmd setup
```


```{r load_packages, include=FALSE, echo = FALSE }
#LOAD PACKAGES -- This will check if a package is installed and if not, install it
ipak <- function(pkg){
  new.pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]
  if (length(new.pkg)) 
    install.packages(new.pkg, dependencies = TRUE, repos = "http://cran.us.r-project.org")
  sapply(pkg, require, character.only = TRUE)
}
#must set up the CRAN location url
packages <- c("foreign", "lessR", "rmarkdown", "tidyverse", "gt", "glue", "AER", "psych", "Hmisc", "ggplot2", "relaimpo")
#have to set the packages
ipak(packages)
```
### Packages used in this analysis
```{r, echo = FALSE}
packages
```




```{r load_data}
# load the California Schools Dataset and give the dataset a shorter name
data(CASchools)
cas <- CASchools
# Convert grade to numeric
# table(cas$grades)
cas$gradesN <- cas$grades == "KK-08"
# Get the set of numeric variables
v <- setdiff(names(cas), c("district", "school", 
    			"county", "grades"))
```

### Q 1 What does the CASchools dataset involve?
Quoting the help (i.e., `?CASchools`), the data is "from all 420 K-6 and K-8 districts in California with data available for 1998 and 1999" and the variables are:

    * district: character. District code.
    * school: character. School name.
    * county: factor indicating county.
    * grades: factor indicating grade span of district.
    * students: Total enrollment.
    * teachers: Number of teachers.
    * calworks: Percent qualifying for CalWorks (income assistance).
    * lunch: Percent qualifying for reduced-price lunch.
    * computer: Number of computers.
    * expenditure: Expenditure per student.
    * income: District average income (in USD 1,000).
    * english: Percent of English learners.
    * read: Average reading score.
    * math: Average math score.

Let's look at the basic structure of the data frame. i.e., the number of observations and the types of values:

```{r}
str(cas)
# Hmisc::describe(cas) # For more extensive summary statistics
```



### Q. 2  To what extent does expenditure per student vary?
```{r cas2, message=FALSE, warning=FALSE}
qplot(expenditure, data = cas) + xlim(0, 8000) + 
		xlab("Money spent per student ($)") +
		ylab("Count of schools")
round(t(psych::describe(cas$expenditure)), 1)
```

The greatest expenditure per student is around double that of the least expenditure  per student.


### Q. 3a  What predicts expenditure per student?
```{r}
# Compute and format set of correlations
corExp <- cor(cas["expenditure"], 
		cas[setdiff(v, "expenditure")])
corExp <- round(t(corExp),2)
corExp[order(corExp[,1], decreasing = TRUE), , 
		drop = FALSE]
```

More is spent per student in schools :

1. where people with greater incomes live
2. reading scores are higher
3. that are K-6


### Q. 4  what is the relationship between district level maths and reading scores?
```{r cas4, message=FALSE}
ggplot(cas, aes(read, math)) + geom_point() + 
		geom_smooth()
```

At the district level, the correlation is very strong (r = The correlation is `r round(cor(cas$read, cas$math), 2)`). From prior experience I'd expect correlations at the individual-level in the .3 to .6 range.  Thus, these results are consistent with group-level relationships  being much larger than individual-level relationships.

### Q. 5 What is the relationship between maths and reading after partialling out other effects?


```{r}
# command has strange syntax requiring column numbers rather than variable names
partial.r(cas[v], 
          c(which(names(cas[v]) == "read"), which(names(cas[v]) == "math")), 
          which(!names(cas[v]) %in% c("read", "math"))
          )
```

The partial correlation is still very strong but is substantially reduced.


### Q. 6 What fraction of a computer does each student have?
```{r}
cas$compstud <- cas$computer / cas$students
describe(cas$compstud)
qplot(compstud, data = cas)
```

The mean number of computers per student is `r round(mean(cas$compstud), 3)`.


### Q. 7 What is a good model of the combined effect of other variables on academic performance (i.e., math and read)?
```{r cas7}
# Examine correlations between variables
psych::pairs.panels(cas[v])
```

`pairs.panels` shows correlations in the upper triangle, scatterplots in the lower triangle, and variable names and distributions on the main diagonal.
After examining the plot several ideas emerge.

```{r cas7.transformation, tidy=FALSE}
# (a) students is a count and could be log transformed
cas$studentsLog <- log(cas$students)
# (b) teachers is not the variable of interest:
#	it is the number of students per teacher
cas$studteach <- cas$students /cas$teachers
# (c) computers is not the variable of interest:
#  it is the ratio of computers to students
# table(cas$computer==0) 
# Note some schools have no computers so ratio would be problematic.
# Take percentage of a computer instead
cas$compstud <- cas$computer / cas$students 
# (d) math and reading are correlated highly, reduce to one variable
cas$performance <- as.numeric(
		scale(scale(cas$read) + scale(cas$math)))
```
Normally, I'd add all these transformations to an initial data transformation file that I call in the first block, but for the sake of the narrative, I'll leave them here.

Let's examine correlations between predictors and outcome.
```{r}
m1cor <- cor(cas$performance, 
		cas[c("studentsLog", "studteach",	"calworks",  
						"lunch", "compstud", "income", 
						"expenditure", "gradesN")])
t(round(m1cor, 2))
```


Let's examine the multiple regression.
```{r}
m1 <- lm(performance ~ studentsLog + studteach + 
				calworks + lunch + compstud
				+ income + expenditure + grades, data = cas)		
summary(m1)
```
And some indicators of predictor relative importance.
```{r}
# calc.relimp from relaimpo package.
(m1relaimpo <- calc.relimp(m1,	type="lmg",	rela=TRUE))
```

Thus, we can conclude that:

1. Income and indicators of income (e.g., low levels of lunch vouchers) are the two main predictors. Thus, schools with greater average income tend to have better student performance.
2. Schools with more computers per student have better student performance.
3. Schools with fewer students per teacher have better student performance.

For more information about relative importance and the `relaimpo` package measures check out [Ulrike Grömping's website](http://prof.beuth-hochschule.de/groemping/relaimpo/).
Of course this is all observational data with the usual caveats regarding causal interpretation.

## Now, let's look at some weird stuff.
